---
title: 'ICTWS 2023: R Workshop'
author: "Rob Ritson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# GIS Exercise

## Background
Intro text...

This exercise assumes basic familiarity with foundational geographic information science concepts:
- Projections and Coordinate Reference Systems
- Vector (i.e. points and polygons) and raster data types

## Package Installation and Set-up

We need to the 'sf' package for working with vector data (points and polygons), 'terra' for working with raster data, 'tidyterra' extends 'ggplot2' to work with 'terra' objects, 'maptiles' provides access to basemap imagery, 'move' and 'ctmm' are packages used for the analysis of wildlife tracking data and include real-world example tracking datasets we can practice with

```{r install and load,eval=FALSE}
# Install
install.packages('sf')
install.packages('terra')
install.packages('tidyterra')
install.packages('maptiles')
install.packages('move') #optional
install.packages('ctmm') #optional
```
```{r}
# Load
require(sf)
require(terra)
require(ggplot2)
require(tidyterra)
require(maptiles)
require(move) #optional
require(ctmm) #optional
```

## Create Simple Data
First, let's use `sf` to create a simple point feature.
```{r create a point}
# Create simple point feature
point1 <- st_point(c(1, 3))
point1
plot(point1,col='red',pch=16,cex=2,main="A Simple Point Feature")
```
Next, we are going to create a simple polygon. 
```{r create a polygon}
# Create simple polygon feature
poly1 <- st_polygon(list(rbind(c(0, 0), c(5, 0), c(5, 5), c(0, 5), c(0, 0))))
poly1
plot(poly1,col='blue',border='red',main="A Simple Polygon Feature")
```
Notice that we need 5 coordinates for a four sided shape. This is because we need to be sure to close the shape so it is valid. The first and last coordiantes of your polygon should always be the same.

Shifting to the 'terra' package, lets create a simple raster with 100 cells and give the cells sequential values from 0 to 5.
```{r create a raster}
# Create simple raster
raster1 <- rast(nrows = 10, ncols = 10)
values(raster1) <- seq(0,5,1)
raster1
plot(raster1,main="A Simple Raster")
```
You can see that the colors of the raster cells repeat in sequences of six based on the assigned values.

Next, lets extract the raster cell value to the point we created. Since we created the point with `sf`, we need to first coerce into a format usable by `terra`.
```{r Extract Value}
# Convert 'sf' XY object to 'terra' SpatVector object
vect_pt <- terra::vect(point1)

# Extract raster value
extracted_value <- extract(raster1, vect_pt)
extracted_value
```
"ID" corresponds with the point number and "lyr.1" corresponds with the layer name of the raster we created.

We can examine whether the extracted value was correct by plotting the point over the raster to visually inspect the cell value.
```{r plot point and raster}
# Is the extracted value correct?
terra::plot(raster1,main="Raster with Point")
terra::plot(vect_pt,cex=1.25,col='blue',add=T)
```
Do they match up?

## Simple Map
Now that you've been introduced to 'sf' and 'terra', we can incorporate them with what we've learned in 'ggplot2'. Let's begin by plotting the 'sf' point with 'ggplot2'.
```{r first point map}
# Plot point
ggplot(point1) + geom_sf()
```
Compare to differences of 'geom_sf' to 'geom_point'...

Next, try plotting the polygon and the point together with 'ggplot2'.
```{r first polygon map}
# Plot polygon
ggplot() + 
  geom_sf(data=poly1,fill='red',color='blue',linewidth=2) +
  geom_sf(data=point1,color='blue',size=2)
```

Up until now, coordinate reference hasn't been a big deal. However, its always important to make sure the projections of all your layers are identical before trying to plot them together.
```{r first raster map}
#Set CRS
## Convert 'sf' objects to 'terra'
poly1_vect <- terra::vect(poly1)
point1_vect <- terra::vect(point1)

#insert code before and after to test projections of all layers....
## Set CRS
terra::crs(poly1_vect) <- terra::crs(raster1)
terra::crs(point1_vect) <- terra::crs(raster1)

#Clip raster extent to polygon
ext <- terra::ext(poly1_vect)*15
raster2 <- terra::crop(raster1,ext)

# Plot raster
ggplot() + 
  geom_spatraster(data = raster2) +
  geom_sf(data=poly1_vect,color='red',fill='yellow',linewidth=1) +
  geom_sf(data=point1_vect,color='purple',size=1.5)
```
Our polygon is very small, so we want to also adjust the extent of the raster before we plot it so we can actually see the polygon. 

You've just made your first simple map with 'ggplot2'!

### chatGPT's tips to becoming a data literate wildlife biologist
* Learn the basics of statistics: Wildlife biology often involves the analysis of large amounts of data, and a strong understanding of statistics is essential for making sense of this data. Start by learning the basics of probability and statistics, such as descriptive statistics, probability distributions, and inferential statistics.

* Learn data manipulation and visualization: Wildlife biologists often need to work with large datasets, and it's important to be able to manipulate and visualize this data effectively. Familiarize yourself with tools such as R or Python for data manipulation and visualization, and learn how to use packages such as "dplyr" or "pandas" for data wrangling.

* Learn GIS: Geographic information systems (GIS) are widely used in wildlife biology to map and analyze spatial data. Learn how to use GIS software such as ArcGIS or QGIS and become familiar with concepts such as projections, coordinate systems, and spatial data formats. Additionally, learn how to use R packages such as 'sf' and 'terra' for GIS. Understanding how to map and analyze spatial data will be extremely useful in a wildlife biology career.

## Challenge
expand this and elaborate... include packages that have these test shapes and rasters as internal files...
```{r,eval=F}
lux <- sf::st_read(system.file("ex/lux.shp", package="terra"))

head(lux)

sf::st_crs(lux)

lux_areas <-sf::st_area(lux) # square meters
lux_areas_km2 <- as.numeric(lux_areas / 1000000) # square meters per square kilometer
lux_areas_km2

## OR ##
units::set_units(lux_areas,m^2) # square meters
units::set_units(lux_areas,km^2) # square kilometers
units::set_units(lux_areas,ha) # hectares
units::set_units(lux_areas,acres) # acres
units::set_units(lux_areas,mi^2) # square miles
# # # # # # # # #

lux_rast <- terra::vect(lux)

elev <- terra::rast(system.file("ex/elev.tif", package="terra"))

terra::crs(lux_rast) == terra::crs(elev)

lux_proj <- terra::project(lux_rast,elev)

terra::crs(lux_proj) == terra::crs(elev)

require(dplyr)
lux_elev <- terra::extract(elev,lux,method = "simple") %>%
  dplyr::group_by(ID) %>%
  dplyr::summarise(Mean = mean(elevation,na.rm=T)) %>%
  as.data.frame(.)
lux_elev

require(ggplot2)

lux_shp <- lux %>%
  dplyr::mutate(NAME = NAME_2) %>%
  dplyr::select(NAME,POP,geometry) %>%
  dplyr::mutate(AREA_km2 = lux_areas_km2,
                ELEVATION_Mean = lux_elev$Mean)
lux_shp

ggplot(data = lux_shp) +
  geom_sf(aes(fill=AREA_km2))

ggplot(data = lux_shp) +
  geom_sf(aes(fill=ELEVATION_Mean))

ggplot(data = lux_shp) +
  geom_sf(aes(fill=POP))

###
## Calculate Hillshade
slp <- terra::terrain(elev,"slope",unit="radians")
asp <- terra::terrain(elev,"aspect",unit="radians")
hillshade <- terra::shade(slp,asp,10, 340)

terra::plot(hillshade,col=grey(0:100/100), legend=FALSE, mar=c(2,2,1,4))

install.packages('tidyterra')
require(tidyterra)
require(ggplot2)
ggplot(lux_shp) +
  geom_spatraster(data=hillshade) +
  geom_spatvector(fill=NA) +
  geom_spatvector_text(aes(label=NAME),color ="grey90") +
  coord_sf(crs = 3857)
  
install.packages('ggnewscale')
require(ggnewscale)
ggplot() +
  geom_spatraster(data = hillshade,show.legend = F,alpha=1) +
  scale_fill_gradientn(colours = grey(0:100 / 100), na.value = NA) +
  new_scale_fill() +
  geom_spatvector(data=lux_shp,aes(fill=POP),alpha=0.4,color="black") +
  scale_fill_whitebox_c(palette = "soft") +
  geom_spatvector_text(data=lux_shp,aes(label=NAME),color ="grey90") +
  coord_sf(crs = 3857)


```


### chatGPT's tips for new R users
* Start by learning the basics: Before diving into complex analyses, it's important to understand the basics of R, such as data types, data structures, and basic operations. This will make it easier to understand more advanced concepts and coding practices.

* Use the built-in help function: R has an extensive help system that can be accessed by typing "?function_name" in the console. This will give you information about the function, its arguments, and examples of how to use it.

* Learn the data manipulation packages: R has several powerful packages for data manipulation such as "dplyr", "tidyr", and "data.table" which are very useful to manipulate data.

* Keep your code organized: As you work on larger projects, it's important to keep your code organized. Use meaningful variable and function names, comment your code, and use indentation to make it easy to read and understand.

* Use the community: R has a large and active community of users. You can find answers to your questions on online forums, Stack Overflow, and other resources. Additionally, there are many R packages available for various analyses, which can make your work much easier.

* Practice, practice, practice: The best way to learn R is by practicing. Start with small projects and work your way up to more complex analyses. The more you practice, the more comfortable you will become with the language.


## Wildlife Location Data Challenge
Expand and elaborate on this exercise....
```{r}
install.packages('move')
require(move)

fisher_raw <- move::fishers
fisher_locs <- terra::vect(fisher_raw)
fisher_locs
head(fisher_locs)
terra::plot(fisher_locs)

library(ctmm)
data("buffalo") #"coati","jaguar","pelican"
head(buffalo$Cilla)
df <- data.frame(buffalo$Cilla)
fisher_locs <- sf::st_as_sf(df,
                        coords = c("x","y"),
                        crs = buffalo$Cilla@info$projection) %>%
  terra::vect(.)

## Map
require(maptiles)
require(dplyr)
#Get background map tile
  tile_zoom5 <- get_tiles(fisher_locs, provider = "Esri.WorldStreetMap", crop=F, zoom=5, cachedir = ".") %>%
    terra::project(.,terra::crs(fisher_locs))
  
  tile_zoom8 <- get_tiles(fisher_locs, provider = "Esri.WorldStreetMap", crop=F, zoom=8, cachedir = ".") %>%
    terra::project(.,terra::crs(fisher_locs))

  tile_zoom10 <- get_tiles(fisher_locs, provider = "Esri.WorldImagery", crop=F, zoom=15, cachedir = ".") %>%
    terra::project(.,terra::crs(fisher_locs))
  #Create Full Map
  map_z5 <- ggplot() + 
    geom_spatraster_rgb(data = tile_zoom5, maxcell = Inf) +
    geom_sf(data = fisher_locs, fill = NA, color = "#000000",) +
    coord_sf(datum = pull_crs(fisher_locs)) + theme_minimal()
  
  map_z8 <- ggplot() + 
    geom_spatraster_rgb(data = tile_zoom8, maxcell = Inf) +
    geom_sf(data = fisher_locs, fill = NA, color = "#000000",) +
    coord_sf(datum = pull_crs(fisher_locs)) + theme_minimal()
  
  map_z10 <- ggplot() + 
    geom_spatraster_rgb(data = tile_zoom10, maxcell = Inf) +
    geom_sf(data = fisher_locs, fill = NA, color = "yellow",) +
    coord_sf(datum = pull_crs(fisher_locs)) + theme_minimal()

map_z5
map_z8  
map_z10
```
A simple field map!
